A fault-tolerant distributed infrastructure for massively multiplayer online games

### 2011-07-14 textos/phd-prospectus/prospectus.tex

\abstract {
Massively multiplayer online games (MMOGs) have become an important genre of online entertainment in the last decade. In these games, many thousands of participants play simultaneously with one another in the same match. This kind of application has been traditionally supported by powerful and expensive central infrastructures with considerable cpu power and efficient Internet connections. This work proposes a geographically distributed infrastructure to support MMOGs formed by nodes which help serve the game to the players. The most important features that must be provided are: consistency (since different players should have the sense of one single shared virtual environment), timeliness (as these games are usually real-time applications) and fault tolerance (as some of the participating nodes may crash and become unavailable).
}

@@

MMOGs emerged in the last decade as a new trend because of the decrease of the domestic Internet connection cost and the increase of its average bandwidth. This kind of games have become very popular because they allow the online interaction of a massive number of players at the same time. In the most succesful cases, such as World of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and EVE Online \cite{eveonline}, for example, tens of thousands of players are supported simultaneously \cite{chen2006pch}. In these games, the players can interact with one another in a virtual environment.

@@

In such peer-to-peer approach, the players would need to agree upon the outcome of the actions, in order for their view to be consistent. To reduce the traffic between peers, the state updates could be sent only to whom they are relevant. To define what is relevant for each player, the virtual environment could be divided into regions \cite{schiele2007rpp}, so that players whose avatars were in a given region could form a small peer-to-peer group. This way, instead of all peers having to agree upon the outcome of every action, only the peers in a given group would have to agree upon actions which were relevant to that group. Unfortunately, nevertheless, the fully decentralized approach would imply a heavy communication cost on the players, specially when running some agreement protocol. This might not only degrade the quality of the game, but also increase the requirements imposed on the players' computational resources, therefore reducing the number of potential participants.

@@

We consider here persistent-state real-time games with virtual environments where each player controls an avatar -- a virtual object that represents him -- by issuing commands to it. Through the avatar, the player interacts with the virtual world and with other objects present in it, such as avatars from other players. The basic operation of the game infrastructure must be as follows: when a player connects to the server, it must send the current state of the player's avatar and of the surroundings of its current location; after that, this player keeps receiving periodic state updates for the objects which are relevant to him -- usually, objects are considered relevant to a player when he can see them \cite{bezerra2008a3,ahmed2008dai,minson2005aim}. When a player issues a command to the avatar (e.g., move from one place to another, pick up an object, attack another player's avatar), it is sent as a message to the server, which processes it and decides its outcome, probably changing the state of the game, which is then broadcast to all involved players.

@@

\begin{itemize}
	\item \textbf{Simulation} of the game, that is, the processing of commands sent by players to their avatars, calculating the corresponding outcomes, which may result in changes to the game state;
	\item \textbf{Communication}, related to broadcasting the current game state to the different players, so that their view is consistent with the changes that might have happened;
	\item \textbf{Storage} of the game state, which is the combination of the states of all objects in the virtual environment.
\end{itemize}

It is important to note that in real-time games (as opposed to turn-based games), due to the possibly fast-paced interaction, it is critical to provide \textbf{timeliness} for the delivery of state update messages to the players, so that they have the illusion of a shared environment. For the same reason, it is necessary to provide \textbf{consistency} for the game state between each pair of players interacting with one another, which means that the game state they perceive must be as similar as possible. Although it is very hard to guarantee that every player will share the exact same view, it is important for their commands to be processed as if there was a single, strongly consistent game state, which is the one stored at the server.

@@

The main purpose of the system is to reduce or, ideally, eliminate the necessity of a central server for MMOGs. The approach we propose to achieve this is based on principles similar to those of volunteer computing, where a potentially large set of contributors make their computers available to perform tasks requested by some remote entity. With such a system, although each of the nodes is probably much less powerful than a traditional MMOG server, when working together they can sum up the cpu power and the network bandwidth necessary to host one of these games.

However, several questions arise when dealing with a potentially large, geographically distributed server system composed of voluntary nodes. First, as the nodes are volunteer, there is no guarantee that they will be available whenever they are needed, neither is there any guarantee regarding the amount of resources that each of these nodes makes available for the system -- be it cpu power, network bandwidth or storage capability. Group communication could be used, but it would have limited efficiency, since network level multicast is not widely available on the Internet.

Finally, considering the potentially large number of nodes in this system, it is very likely that some of them present failures, for what we are assuming the crash-stop failure model -- therefore, we assume the possibility of nodes crashing and network links becoming unavailable, but we consider that the servers will not present Byzantine behavior. The services provided by the system must continue even in the presence of such failures. Not only must the distribution itself be transparent to the players, but also the existence of failures and their countermeasures should not be noticeable by them.

In Section \ref{sec:statepart}, we describe how the game state will be distributed among the different server nodes and, in \mbox{Section \ref{sec:consprot}}, a preliminary consistency management protocol, based on state machine replication, is presented. We also introduce, in Section \ref{sec:optmcast}, our group communication protocol on top of which our consistency management is developed.

@@

In order to split a game server into several smaller units, executed by many volunteer nodes, we divide the virtual environment in regions, each of which managed by a different server. Different approaches can be found in the literature, such as dividing the space intousing fixed-size cells and grouping them to form regions \cite{devleeschauwer2005dma}, or using binary space partition trees \cite{bezerra2009fgl}. To tolerate failures, each server has several replicas, forming a server group.

The state partitioning is performed based on which objects are located in each region. The state of an object is managed, then, by the server group assigned to the region where that object is located. This way, the locality of data is explored: objects close to each other are more likely to interact and, as they are located on the same server group, their interaction should be faster. Besides, by exploring the locality of the game data (state of the objects), the inter-group communication overhead is reduced.

By having multiple server replicas in each region, the cost of broadcasting state update messages to the players is also reduced. Since all the replicas in a group are supposed to have an identical state, each one of them may send state updates to a different subset of the players located in that region. As the number of players may be as large as many thousands, this would probably be the most costly operation for the system and distributing it would have a significant impact on the system scalability.

@@

As mentioned before, the players interact by issuing commands and having them processed by servers. Servers calculate the resulting game state and send it to the players. Every command can alter the state of the game, possibly based on a previous state. We assume that commands lead to a deterministic sequence of game states. In other words, if the game starts in the state $S$, there is only one possible state $S'$ to be reached after a sequence $C=\{c_{1}, c_{2}, ..., c_{n}\}$ of commands.

The game state could then be modeled as a state machine, which would be replicated across the different server nodes, as proposed in \cite{lamport1978ird,schneider1990ift,lampson1996hbh}, and each command would be responsible for a state transition. One way of implementing the distributed server system would be by creating such a state machine and forwarding all commands to every server node, in the same (total) order. As the state transitions are deterministic, every server would be able to calculate the correct game state independently. It might be the case that some servers could still be processing an older command, while others might have already processed newer inputs. Although this might lead to an old state, it would never lead to an incorrect state. To check which one is the newest, it would be enough to have a version number considering, for instance, the number of commands processed, since the commands arrive in the same order at the different servers. We would, then, have a fully replicated server system.

@@

Instead of forwarding each command to every server, now only the servers that manage some object in the target set of a command would process it. The problem with this scheme is that, in order to execute some command, a group might need the previous state of an object managed by some other group. To solve this, a possibility would be to have the groups requesting the states they need from other groups. Another solution, probably more latency-efficient, would be every group proactively sending the new states of the objects managed by it to other groups which might need them.

Sending new states to other groups proactively based on the possibility that they might need such states, however, may lead to all groups sending to each other the state changes of every object, degenerating to a fully replicated approach. Nevertheless, we can explore the fact that, in games, objects interact only with other objects situated nearby: we can define the regions of the virtual environment in a way such that only objects either in the same or in neighboring regions would be able to interact -- two regions are neighbors if, and only if, they share a border. As commands are executed by avatars, which are simply objects controlled by players, the result will be that, for each pair of objects in the target set of any command, these two objects will be either located in the same or in neighbor regions. This way, only groups managing neighbor regions would ever execute the same command and each group would never need a state from another group which is not its neighbor.

Although unlikely, it is possible to have a command sequence $C = \{c_1,c_2,\ldots,c_n\}$ where (1) every $c_i$ depends on the state resulting from $c_{i-1}$, (2) not every $c_i$ is processed by the same group and (3) not all of these groups manage neighbor regions. Nevertheless, the servers can detect such dependencies and forward the object states to the servers that need them.% In the worst case, every group but one -- there is no dependency cycle -- would be waiting for another group, so the time needed to process a command would be proportional to the number of groups.

% Figure \ref{fig:consistency} illustrates two avatars situated in regions managed by different server groups, but still close enough to interact. Each command issued by their players has to be forwarded to the groups that manage some object in its target set, which is done by multicast.
% 
% \begin{figure}[!h]
%   \centering
%   \includegraphics[width=0.8\linewidth]{images/consistency}
%   \caption{Multicasting a command to a group which manages an object in that command's target set}
%   \label{fig:consistency}
% \end{figure}

@@

To propagate commands to servers we are employing a multicast communicastion protocol. 
%As the servers are organized in groups, and every command is sent to one or more of such groups, we can use a multicast protocol. Also, as 
As commands seen by different servers must be handled in the same order, we need total order, so that the state machine replication approach can be used. Besides, the commands sent by each player should be executed in the same order in which they have been issued; therefore, the multicast protocol should provide FIFO ordering.

However, the inter-group communication overhead should be minimized, so that the distributed server would be more scalable. Non-genuine multicast protocols \cite{schiper2009gng} require control messages to be exchanged between groups, even when there is no application level message to be sent. Genuine multicast algorithms minimize the number of messages exchanged, but they impose a greater delivery latecy \cite{schiper2008ica}.

We devised a multicast protocol that, although not genuine, allows the system to increase its scale without getting flooded with control messages. We call it \emph{quasi-genuine} because every group keeps sending messages to (and receiving from) other groups, but only a few of them. This is because the protocol takes into account the groups with which a group may exchange messages. The information concerning which other groups a given group can communicate with must be given by the application.

The advantage of a quasi-genuine multicast protocol is that it would be a middle ground between genuine and non-genuine protocols, inheriting some advantages from both, namely, improved scalability and lower message delivery time, at the cost of restricting which groups can communicate with each other. Although the system can be reconfigured to change the group pairs that are able to exchange messages, such reconfiguration will have some cost, so this multicast protocol is more suitable for applications where this would not be frequently required.

This communication primitive can be used by our consistency management algorithm because every server group only has to forward commands and states to a specific set of other groups, which are the ones managing neighbor regions; they are known as soon as the game partitioning has been performed. Changes to the partioning should be made rarely, as it would incur in some players migrating between servers. Besides, changing the regions does not necessary change which groups can communicate with each other. Therefore, our quasi-genuine multicast protocol is well suited for the consistency management protocol we are devising.

To further decrease the message delivery time, an optimistic delivery is also being considered. By assigning timestamps based on the local clocks of the servers -- which are supposed to be more or less synchronized -- and by every server waiting long enough for messages to arrive from other servers before performing the delivery, the final total message order will probably match the timestamp order. This way, messages could be delivered optimistically in only one communication step, although this delivery would have to be eventually confirmed by a conservative one.
%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We are currently working towards the validation of the quasi-genuine multicast and the game-state consistency management protocols, which require both formal proofs, regarding correctness, and experimental results to demonstrate their performance. Another direction that we intend to follow is to consider not only crash-stop, but also Byzantine faults. By tolerating these faults, the server system would be resistant to many kinds of cheating \cite{yan2005scc}, which can also be seen as Byzantine behavior.
%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%
\bibliographystyle{abbrv}
\bibliography{references}

>>>>

\abstract {
Massively multiplayer online games (MMOGs) have become an important genre of online entertainment in the last decade. In these games, many thousands of participants play simultaneously with one another in the same match. This kind of application has been traditionally supported by powerful and expensive central infrastructures with considerable cpu power and efficient Internet connections. This work proposes a geographically distributed infrastructure to support MMOGs, formed by nodes that help serve the game to the players. The most important features that must be provided are: consistency, since different players should have the sense of one single shared virtual environment, timeliness, as these games are usually real-time applications, and fault tolerance, as some of the participating nodes may crash and become unavailable.
}

@@

MMOGs emerged in the last decade as a new trend because of the decrease of the domestic Internet connection cost and the increase of its average bandwidth. This kind of games have become very popular because they allow the online interaction of a massive number of players at the same time. In the most successful cases, such as World of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and EVE Online \cite{eveonline}, for example, tens of thousands of players are supported simultaneously \cite{chen2006pch}. In these games, the players can interact with one another in a virtual environment.

@@

In such fully decentralized approach, the players would need to agree upon the outcome of the actions, in order for their view to be consistent. To reduce the traffic between peers, each state update could be sent only to players to whom it is relevant. To define what is relevant to each player, the virtual environment could be divided into regions \cite{schiele2007rpp}, so that players whose avatars were in a given region could form a small peer-to-peer group. This way, instead of all peers having to agree upon the outcome of every action, only the peers in a given group would have to agree upon actions which were relevant to that group. Unfortunately, this approach would still imply a heavy communication cost on the players, specially when running some agreement protocol. This might not only degrade the quality of the game, but also increase the requirements imposed on the players' computational resources, therefore reducing the number of potential participants.

@@

We consider here persistent state real-time games with virtual environments where each player controls an avatar -- a virtual object that represents him -- by issuing commands to it. Through the avatar, the player interacts with the virtual world and with other objects present in it, such as avatars from other players. The basic operation of the game infrastructure must be as follows: when a player connects to the server, the server must send the current state of the player's avatar and of the surroundings of its current location; after that, this player keeps receiving periodic state updates for the objects which are relevant to him -- usually, objects are considered relevant to a player when he can see them \cite{bezerra2008a3,ahmed2008dai,minson2005aim}. When a player issues a command to the avatar (e.g., move from one place to another, pick up an object, attack another player's avatar), it is sent as a message to the server, which processes it and decides its outcome, probably changing the state of the game, which is then broadcast to all involved players.

@@

\begin{itemize}
	\item \textbf{Simulation} of the game world, that is, the processing of commands sent by players to their avatars, calculating the corresponding outcomes, which may result in changes to the game state;
	\item \textbf{Communication} of the current game state to the different players, so that their view is consistent with the changes that might have happened;
	\item \textbf{Storage} of the game state, which is the combination of the states of all objects in the virtual environment.
\end{itemize}

It is important to note that in real-time games, as opposed to turn-based games, due to the possibility of fast-paced interactions, it is critical to provide \textbf{timeliness} for the delivery of state-update messages to the players, so that they have the illusion of a shared environment. For the same reason, it is necessary to provide \textbf{consistency} for the game state between each pair of players interacting with one another, which means that the game state they perceive must be as similar as possible. Although it is hard to guarantee that every player will share the exact same view, it is important for their commands to be processed as if there was a single, strongly consistent game state, which is the one stored at the server.

@@

The main purpose of the infrastructure is to reduce or, ideally, eliminate the necessity of a central server for MMOGs. The approach we propose to achieve this is based on principles similar to those of volunteer computing, where a potentially large set of contributors make their computers available to perform tasks requested by some remote entity. With such a system, although each of the nodes is probably much less powerful than a traditional MMOG server, when working together they can sum up the cpu power and the network bandwidth necessary to host one of these games.

However, several questions arise when dealing with a potentially large, geographically distributed server system composed of voluntary nodes. First, as the nodes are volunteer, there is no guarantee that they will be available whenever they are needed, neither is there any guarantee regarding the amount of resources that each of these nodes makes available for the system -- be it cpu power, network bandwidth or storage capability.% Group communication could be used, but it would have limited efficiency, since network level multicast is not widely available on the Internet.

Finally, considering the potentially large number of nodes in this system, it is likely that some of them present failures. For simplicity, we assume the crash-stop failure model, where servers do not present Byzantine behavior. The services provided by the system must continue even in the presence of failures. Not only must the distribution itself be transparent to the players, but also the existence of failures and their countermeasures should not be noticeable by them.

In Section \ref{sec:statepart}, we describe how the game state will be distributed among the different server nodes and, in \mbox{Section \ref{sec:consprot}}, we present a preliminary consistency management protocol, based on state machine replication. We also introduce, in Section \ref{sec:optmcast}, our group communication protocol on top of which our consistency management will be developed.

@@

In order to split a game server into several smaller units, executed by many volunteer nodes, we divide the virtual environment in regions, each of which managed by a different server. Different approaches can be found in the literature, such as dividing the space using fixed-size cells and grouping them to form regions \cite{devleeschauwer2005dma}, or using binary space partitioning trees \cite{bezerra2009fgl}. To tolerate failures, each server has several replicas, forming a server group.

The state partitioning is performed based on which objects are located in each region. The state of an object is managed, then, by the server group assigned to the region where that object is located. This approach exploits data locality: objects close to each other are more likely to interact and, as they are located on the same server group, their interaction should be faster. Besides, by exploiting the locality of the game data (state of the objects), the inter-group communication overhead is reduced.

As the number of players may be as large as many thousands, broadcasting state-update messages to them will probably be the most costly operation for the server system. By having multiple server replicas in each region, this cost will also be distributed: since all the replicas in a group are supposed to have an identical state, each one of them may send state updates to a different subset of the players whose avatars are located in that group's region.

@@

As mentioned before, players interact by issuing commands and having them processed by servers. Servers calculate the resulting game state and send it to the players. Every command can alter the state of the game, possibly based on a previous state. We assume that commands lead to a deterministic sequence of game states. In other words, if the game starts in the state $S$, there is only one possible state $S'$ to be reached after a sequence $C=\{c_{1}, c_{2}, ..., c_{n}\}$ of commands is executed.

The game state could be modeled as a state machine, which would be replicated across the different server nodes, as proposed in \cite{lamport1978ird,schneider1990ift,lampson1996hbh}, and each command would be responsible for a state transition. One way of implementing the distributed server system would be by creating such a state machine and forwarding all commands to every server node, in the same (total) order. As the state transitions are deterministic, every server would be able to calculate the correct game state independently. %It might be the case that some servers could still be processing an older command, while others might have already processed newer inputs. Although this might lead to an old state, it would never lead to an incorrect state. To check which one is the newest, it would be enough to have a version number considering, for instance, the number of commands processed, since the commands arrive in the same order at the different servers. 
We would, then, have a fully replicated server system.

@@

% \begin{figure}[!h]
%   \centering
%   \includegraphics[width=0.8\linewidth]{images/consistency}
%   \caption{Each command is processed by every group that manages some object in that command's target set}
%   \label{fig:consistency}
% \end{figure}

Instead of forwarding each command to every server, now only the servers that manage some object in the target set of a command would process it. % (Figure \ref{fig:consistency}).
The problem with this scheme is that, in order to execute some command, a group might need the previous state of an object managed by some other group. To solve this, a possibility would be to have the groups requesting the states they need from other groups. Another solution, probably more latency-efficient, would be every group proactively sending the new states of the objects managed by it to other groups which might need them. We intend to look into both alternatives.

%Sending new states to other groups proactively based on the possibility that they might need such states, however, may lead to all groups sending to each other the state changes of every object, degenerating to a fully replicated approach. Nevertheless, we can explore the fact that, in games, objects interact only with other objects situated nearby: we can define the regions of the virtual environment in a way such that only objects either in the same or in neighboring regions would be able to interact -- two regions are neighbors if, and only if, they share a border. As commands are executed by avatars, which are simply objects controlled by players, the result will be that, for each pair of objects in the target set of any command, these two objects will be either located in the same or in neighbor regions. This way, only groups managing neighbor regions would ever execute the same command and each group would never need a state from another group which is not its neighbor.

%Although unlikely, it is possible to have a command sequence $C = \{c_1,c_2,\ldots,c_n\}$ where (1) every $c_i$ depends on the state resulting from $c_{i-1}$, (2) not every $c_i$ is processed by the same group and (3) not all of these groups manage neighbor regions. Nevertheless, the servers can detect such dependencies and forward the object states to the servers that need them.% In the worst case, every group but one -- there is no dependency cycle -- would be waiting for another group, so the time needed to process a command would be proportional to the number of groups.

% Figure \ref{fig:consistency} illustrates two avatars situated in regions managed by different server groups, but still close enough to interact. Each command issued by their players has to be forwarded to the groups that manage some object in its target set, which is done by multicast.
%

@@

To propagate commands to servers we are considering a multicast communication protocol. 
%As the servers are organized in groups, and every command is sent to one or more of such groups, we can use a multicast protocol. Also, as 
As commands seen by different servers must be handled in the same order, we need total order, so that the state machine replication approach can be used. Besides, the commands sent by each player should be executed in the same order in which they have been issued; therefore, the multicast protocol should provide FIFO ordering.

However, the inter-group communication overhead should be minimized, so that the distributed server would be more scalable. Non-genuine multicast protocols \cite{schiper2009gng} require control messages to be exchanged between groups, even when there is no application level message to be sent. Genuine multicast algorithms minimize the number of messages exchanged, but they impose a greater delivery latency \cite{schiper2008ica}.

We devised a multicast protocol that, although not genuine, allows the system to increase its scale without getting flooded with control messages. We call it \emph{quasi-genuine} because every group keeps sending messages to (and receiving from) other groups, but only a few of them. This is because the protocol takes into account 
communication patterns of the application. Such patterns can be inferred from the application semantics and given to the multicast algorithm as a parameter.
%the groups with which a group may exchange messages. The information concerning which other groups a given group can communicate with must be given by the application.

%The advantage of a quasi-genuine multicast protocol is that it would be a middle ground between genuine and non-genuine protocols, inheriting some advantages from both, namely, improved scalability and lower message delivery time, at the cost of restricting which groups can communicate with each other. Although the system can be reconfigured to change the group pairs that are able to exchange messages, such reconfiguration will have some cost, so this multicast protocol is more suitable for applications where this would not be frequently required.

This communication primitive can be used by our consistency management algorithm because every server group only has to forward commands and states to a specific set of other groups, which are the ones managing neighbor regions; they are known as soon as the game partitioning has been performed. Changes to the partitioning should be made rarely, as it would incur in some players migrating between servers. Besides, changing the regions does not necessary change which groups can communicate with each other. Therefore, our quasi-genuine multicast protocol is well suited for the consistency management protocol we are devising.

To further decrease the message delivery time, an optimistic delivery is also being considered. If every server monitors the deviation between its clock and those of other servers, while also estimating the maximum message delivery time, it is possible to deliver messages in only one communication step and in total order. As these estimations may be incorrect, a second, conservative message delivery still must be done to confirm the optimistic one.
%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We are currently working towards the validation of the quasi-genuine multicast and the game state consistency management protocols, which require both correctness proofs and experimental results, to demonstrate their performance. Another direction that we intend to follow is to consider not only crash-stop, but also Byzantine faults. By tolerating these faults, the server system would be resistant to many kinds of cheating \cite{yan2005scc}, which can also be seen as Byzantine behavior.
%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%
\bibliographystyle{ieeetr}
\bibliography{references}
